---
title: 'EDS 223: assignment 2'
author: "Lewis White"
date: "2022-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapters 3](https://geocompr.robinlovelace.net/attr.html), [4](https://geocompr.robinlovelace.net/spatial-operations.html), [5](https://geocompr.robinlovelace.net/geometry-operations.html) of Geocomputation with R by Rovin Lovelace. Each question lists the total number of points. The breakdown of points can be found at the end of each instruction in parentheses. A general grading rubric can be found on the [course website](https://ryoliver.github.io/EDS_223_spatial_analysis/assignments.html).

Please update "author" to list your first and last names and any collaborators (e.g. Ruth Oliver, Friend1, Friend2).

**Due by midnight on Saturday 2022-10-22**

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
# add required packages here
library(sf)
library(tmap)
library(spData)
library(tidyverse)
library(rmapshaper)
library(smoothr)
```

## question 1

##### 5 points

Find the states that:(2.5)\
- belong to the West region\
- have an area below 250,000 square kilometers\
- and greater than 5,000,000 residents in 2015

```{r include=TRUE}
us_states %>%
  filter(REGION == "West") %>%
  mutate(AREA = as.numeric(AREA)) %>%  #the AREA was originally class 'Units' ~ I was able to convert it to numeric without having to remove "km" from the values.
  filter(AREA < 250000) %>%
  filter(total_pop_15 > 5000000) 

print("Washington is the only state that belongs to the West region, has an area under 250,00 square kilometers and had a population over 5,000,000 in 2015.")
```


What was the total population of the US in 2015? (2.5)

```{r include=TRUE}
us_states %>%
  summarise(total_pop_us = sum(total_pop_15))

#314,375,347   EXCEPT ALASKA AND HAWAII (including DC)


print("In 2015, the total population of the contiguous United States was 314,375,347 people. This does not include the population of Hawaii or Alaska, which was not included in the dataset.")
```


## question 2

##### 5 points

Create a new variable named "us_states_stats" by adding variables from "us_states_df" to "us_states". (3)

-   Which function did you use and why? (0.5)
-   Which variable is the key variable in both datasets? (0.5)
-   What is the class of the new object? (0.5)
-   Why does us_states_df have 2 more rows than us_states? (0.5)

```{r include=TRUE}
us_states_stats <- full_join(us_states, us_states_df, by = c("NAME" = "state"))

class(us_states_stats)

print("I used `full_join()` because I wanted to keep all of the information from both datasets.")

print("For the key variable, I used 'NAME' from the us_states dataset and 'state' from the us_states_df dataset. Each of these contains a list of US states.")

print("The class of the new object is an sf data frame.")

print("The us_states data frame only includes the contiguous states (and the District of Columbia). Meanwhile, us_states_df includes Hawaii and Alaska.")
```


## question 3

##### 10 points

Make a map of the percent change in population density between 2010 and 2015 in each state. Map should include a legend that is easily readable. (7)

```{r include=TRUE}
#adding new columns for population density in 2010, population density in 2015, and the percent change in density over this time period.

us_states_pop_density_change <- us_states %>%
  mutate(AREA = as.numeric(AREA)) %>%
  mutate(pop_density_2010 = total_pop_10/AREA) %>%
  mutate(pop_density_2015 = total_pop_15/AREA) %>%
  mutate(pop_density_change = (pop_density_2015/pop_density_2010 - 1) * 100)

tm_shape(us_states_pop_density_change) +
  tm_fill(col = "pop_density_change",
          title = "Percent Population Density Change",
          midpoint = 0) +
  tm_borders() +
  tm_layout(legend.outside = TRUE) +
  tm_layout(title.position = c("left", "top"),
            main.title = "Change in population density from 2010 to 2015")
```

In how many states did population density decrease? (3)

```{r include=TRUE}
us_states_pop_density_change %>%
  filter(pop_density_change < 0) %>%
  arrange()

print("The population density decreased in just two states, Michigan and Rhode Island")
```


## question 4

##### 10 points

How many of New Zealand's high points are in the Canterbury region? (5)

```{r include=TRUE}
#filter to canterbury region in NZ
canterbury <- nz |> 
  filter(Name == "Canterbury")

#subset nz_heights to only include observations that are within the canterbury geometry/region.
c_height <- nz_height[canterbury,]

#find the number of observations
nrow(c_height) #70

print("70 of New Zealand's high points are in the Canterbury region.")
```



Which region has the second highest number of "nz_height" points? And how many does it have? (5)

```{r include=TRUE}
#joining the nz dataset and the nz_height dataset
nz_with_heights <- st_join(nz, nz_height)

#group by the region and summarise the number of high point observations in each region
nz_with_heights %>%
  group_by(Name) %>%
  summarise(count = n()) %>%
  slice_max(count, n = 3) %>% #only show top 3 by count
  arrange(desc(count)) #arrange in descending order


print("The West Coast region has the second highest number of 'nz_height' points. 22 high points are in the West Coast region.")
```


## question 5

##### 15 points

Create a new object representing all of the states the geographically intersect with Colorado.(5)\
Hint: use the "united_states" dataset. The most concise way to do this is with the subsetting method "[".\
Make a map of the resulting states. (2.5)

```{r include=TRUE}
#Creating Colorado object
colorado <- us_states %>%
  filter(NAME == "Colorado")

#subsetting to only include Colorado and states that intersect with it
colorado_intersects <- us_states[colorado, , op = st_intersects]

#making a map of Colorado and its surrounding states
tm_shape(colorado_intersects) +
  tm_polygons(col = "NAME") +
  tm_layout(legend.outside = TRUE) +
  tm_text('NAME')
```

Create another object representing all the objects that touch (have a shared boundary with) Colorado and plot the result.(5)

Hint: remember you can use the argument op = st_intersects and other spatial relations during spatial subsetting operations in base R).

Make a map of the resulting states. (2.5)

```{r}
#just the states that touch Colorado (same as the previous task, but Colorado is not included)
colorado_touch <- us_states[colorado, , op = st_touches]
```

```{r include=TRUE}
#make the map
tm_shape(colorado_touch) +
  tm_polygons(col = "NAME",
              title = "State") +
  tm_layout(legend.outside = TRUE)
```

## question 6

##### 10 points

Generate simplified versions of the "nz" dataset. Experiment with different values of keep (ranging from 0.5 to 0.00005) for **ms_simplify()** and dTolerance (from 100 to 100,000) **st_simplify()**. (5)

Map the results to show how the simplification changes as you change values.(5)

```{r include=TRUE}
#MS SIMPLIFY
ms_simple_0.5 <- ms_simplify(nz, keep = 0.5)

ms_simple_0.005 <- ms_simplify(nz, keep = 0.005)

ms_simple_0.00005 <- ms_simplify(nz, keep = 0.00005)

# ms simple 0.5
keep_0.5 <- tm_shape(ms_simple_0.5) +
  tm_polygons() +
  tm_layout(main.title = "Keep 0.5")

# ms simple 0.005
keep_0.005 <- tm_shape(ms_simple_0.005) +
  tm_polygons() +
  tm_layout(main.title = "Keep 0.005")

# ms simple 0.00005
keep_0.00005 <- tm_shape(ms_simple_0.00005) + 
  tm_polygons() +
  tm_layout(main.title = "Keep 0.00005")


tmap_arrange(keep_0.5, keep_0.005, keep_0.00005)



#ST SIMPLIFY
st_simple_100 <- st_simplify(nz, dTolerance = 100)

st_simple_1000 <- st_simplify(nz, dTolerance = 1000)

st_simple_10000 <- st_simplify(nz, dTolerance = 10000)

#st_simple_100000 <- st_simplify(nz, dTolerance = 100000) #this won't plot


dTolerance_100 <- tm_shape(st_simple_100) +
  tm_polygons() +
  tm_layout(main.title = "dTolerance 100")

dTolerance_1000 <- tm_shape(st_simple_1000) +
  tm_polygons() +
  tm_layout(main.title = "dTolerance 1000")

dTolerance_10000 <- tm_shape(st_simple_10000) +
  tm_polygons() +
  tm_layout(main.title = "dTolerance 10000")

#tm_shape(st_simple_100000) +
  #tm_polygons()

tmap_arrange(dTolerance_100, dTolerance_1000, dTolerance_10000)


print("The map of New Zealand regions gets less precise as keep value approaches 0 (Visvalingam’s algorithm). The map of New Zealand regions gets less precise as dTolerance values increase (Douglas-Peucker Algorithm)")
```


## question 7

##### 10 points

How many points from the "nz_height" dataset are within 100km of the Canterbury region?

```{r include=TRUE}
#filter to just canterbury 
canterbury <- nz %>%
  filter(Name == "Canterbury")

#points within 100,000 meters (or 100km) will have value of 1 (true) assigned to them. Points outside of this distance will have 0 (false) assigned. 

close_to_canterbury <- st_is_within_distance(nz_height, canterbury, dist = 100000) # creates a sgbt (sparse geometry binary predicate) list

summary(lengths(close_to_canterbury) > 0) #counts the number of TRUE and FALSE in the list


print("95 points from the 'nz_height' dataset are within 100km of the Canterbury region.")
```

## question 8

##### 15 points

Find the geographic centroid of the country of New Zealand. How far is it from the geographic centroid of Canterbury?

```{r include=TRUE}
#aggregate NZ (2 ways of doing this)
nz_no_regions <- st_combine(nz) 

nz_no_regions2 <- st_union(nz)

#find the centroid of NZ
nz_centroid <- st_centroid(nz_no_regions2)

#finr the centroid of Canterbury
canterbury_centroid <- st_centroid(canterbury)

#find the distance between the centroids 
st_distance(nz_centroid, canterbury_centroid) #234192.6


print("The geographic centroid of New Zealand is 234.2 kilometers away from the geographic centroid of the Canterbury region")
```