---
title: 'EDS 223: assignment 1'
author: 'Lewis White'
date: "2022-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html). Each question lists the total number of points. The breakdown of points can be found at the end of each instruction in parentheses. A general grading rubric can be found on the [course website](https://ryoliver.github.io/EDS_223_spatial_analysis/assignments.html).


**Due by midnight Saturday 2022-10-08**

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(grid)
library(leaflet)
library(tidyverse)
```

These exercises rely on a new data object based on the **world** and **worldbank_df** datasets from the \*\*spData\* package.

```{r include=TRUE}

#load world dataset, filter to continent of interest, join 

africa = world |>  
  filter(continent == "Africa", !is.na(iso_a2)) |>   #filter to africa
  left_join(worldbank_df, by = "iso_a2") |>  #join worldbank info to world dataset
  dplyr::select(name, subregion, gdpPercap, HDI, pop_growth) |> #select vars of interest
  st_transform("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +lon_0=25")  #adjust type of geom

```

We will also use the **zion** and **nlcd** datasets from the **spDataLarge** package.

```{r include=TRUE}
zion = st_read((system.file("vector/zion.gpkg", package = "spDataLarge")))

#loading the raster data 
data(nlcd, package = "spDataLarge")
force(nlcd)
```

## question 1

##### 10 points

Create a map showing the geographic distribution of the Human Development Index (HDI) across Africa.\
- use both base graphics (hint: use **plot()**) and **tmap**) (4)\
- name two advantages of each based on the experience (3) - name three other mapping packages and an advantage of each (3)

```{r include=TRUE}
#USING NATIVE PLOT
plot(africa["HDI"])

plot(africa) #without indexing, this plots each of the variables


# USING TMAP
tm_shape(africa) + 
  tm_fill(col = "HDI",
          title = "HDI Score")
```

**Advantages of Native `Plot()` function**
1. The `plot()` function was wildly quick and easy to implement. Taking away the HDI index, I was able to generate a plot for each column in the data set by just typing "plot(africa)". If I was doing exploratory data analysis and just wanted a quick look into the data, `plot()` seems like the perfect option. 

2. `plot()` doesn't require an additional R package. This would make it pretty reproducible, allowing anyone with R and access to the data to complete the analysis. 

**Advantages of using tmap**

1. tmap allows for much flexibility in its map making. It's fairly easy to customize maps to one's ideal aesthetic. Also, you can use `tmap_mode()` to create interactive maps, a feature which isn't available using `plot()`. 

2. tmap allows you to work with a wider range of spatial classes.


**3 Other Mapping Packages**
**googleway**
The googleway package allows you to access Google Maps APIs to retrieve data and plot maps. A benefit of this package is its ability to let you work with a ton of data generated from Google Maps. 

**cartogram**
The cartoram package allows you to create customizable cartograms, a map type that represents the mapped area by distorting the geometry of the feature itself (e.g. size of the geometry might depend on another numeric variable).

**globe**
The globe package allows you to plot 3D (and 2D ~ but that isn't quite as exciting) of the earth. 

## question 2

##### 20 points

Extend the **tmap** created for question 1 so the legend has three bins: "high" (HDI above 0.7), "medium" (HDI between 0.55 and 0.7), and "low" (HDI below 0.55). (5)\
- change the legend title (5)\
- change the class labels (5)\
- change the color palette (5)


```{r}
tm_shape(africa) +
  tm_fill(col = "HDI",                                    #color depends on HDI
          breaks = c(0, 0.55, 0.7, Inf),                   #setting up the breaks
          labels = c("Low HDI", "Medium HDI", "High HDI"),  #specifying labels for breaks
          title = "HDI Score",                          
          palette = "viridis") +                          
  tm_layout(
    title = "HDI Ratings Across African Countries",
    title.size = 0.5,
    frame = FALSE,
    scale = 1,
    bg.color = "lightblue")   #background color
  

#save the above plot
HDI_plot <- tm_shape(africa) +
  tm_fill(col = "HDI",
          breaks = c(0, 0.55, 0.7, Inf),
          labels = c("Low HDI", "Medium HDI", "High HDI"),
          title = "HDI Score",
          palette = "viridis") +
  tm_borders() +
  tm_layout(
    title = "HDI Ratings Across African Countries", 
    title.position = c("LEFT", "TOP"),
    title.size = 0.5,
    frame = TRUE,
    scale = 1,
    bg.color = "lightblue")

```


THIS IS THE WAY I ORIGINALLY DID IT BEFORE LEARNING ABOUT THE LABELS OPTION IN TM_FILL
```{r include=TRUE}
# #define the HDI breaks
# breaks <- c(0, 0.55, 0.7, Inf) #how do i make this so each is mutually exclusive?
# 
# 
# #added a new column tospecify low, medium and high HDI
# africa_catigorical_HDI <- africa %>%
#   mutate(hdi_cat = ifelse(HDI < 0.55, "Low",
#                           ifelse(HDI >= 0.55 & HDI < 0.7, "Medium",
#                                  ifelse(HDI >= 0.7, "High",
#                                         ifelse(HDI == NA, "Missing")))))
# 
# #change the hdi column to be a factor and specify the levels
# africa_factor_HDI <- africa_catigorical_HDI %>%
#   mutate(hdi_cat = as.factor(hdi_cat)) %>%
#   mutate(hdi_cat = factor(hdi_cat, levels = c("Low", "Medium", "High")))
#                       
# ## another way of coding this  
#                       #africa_categorical_HDI$HDI_cat <- factor(africa_categorical_HDI,                           #levels = c("Low", "Medium", "High"))
# 
# #make the plot
# tm_shape(africa_factor_HDI) + 
#   tm_fill(col = "hdi_cat",
#           title = "HDI Score",
#           palette = "Pastel1") +
#   tm_borders() +
#   tm_layout(
#     title = "HDI Ratings Across African Countries", ##This doesn't really work
#     title.position = c("RIGHT", "TOP"),
#     title.size = 0.5,
#     frame = FALSE,
#     scale = 1,
#     bg.color = "lightblue")
# 
# #save the chart as a new object
# hdi_plot <- tm_shape(africa_factor_HDI) + 
#   tm_fill(col = "hdi_cat",
#           title = "HDI Score",
#           palette = "Pastel1") +
#   tm_borders() +
#   tm_layout(
#     title = "HDI Ratings Across African Countries", ##This doesn't really work
#     title.position = c("RIGHT", "TOP"),
#     title.size = 0.5,
#     frame = TRUE,
#     scale = 1,
#     bg.color = "lightblue")
```

## question 3

##### 20 points

Represent Africa's subregions on the map. (5)\
- change the color palette (5)\
- change the legend title (5)\
- combine this map with the map from question 2 into a single plot (5)

```{r include=TRUE}

#plot the subregions
tm_shape(africa) + 
  tm_fill(col = "subregion",
          title = "Subregion",
          palette = "Accent") +
  tm_borders() +
  tm_layout(
    title = "African Subregions", 
    title.position = c("LEFT", "TOP"),
    title.size = 0.5,
    frame = FALSE,
    scale = 1,
    bg.color = "lightblue")

#save the map as a new object
subregion_plot <- tm_shape(africa) + 
  tm_fill(col = "subregion",
          title = "Subregion",
          palette = "Accent") +
  tm_borders()  +
  tm_layout(
    title = "African Subregions", 
    title.position = c("LEFT", "TOP"),
    title.size = 0.5,
    frame = TRUE,
    scale = 1,
    bg.color = "lightblue")
```

```{r}
#plot the subregion and hdi plots together in one plot
tmap_arrange(subregion_plot, HDI_plot)
```


## question 4

##### 30 points

Create a land cover map of Zion National Park (5)\
- change the default colors to match your perception of land cover categories (5)\
- move the map legend outside of the map to improve readability (5)\
- add a scale bar and north arrow and change the position of both to improve the maps aesthetics (5)\
- add the park boundaries on top of the land cover map (5)\
- add an inset of Zion's location in the context of the state of Utah (5)\
- hint: an object representing Utah can be subset from the **us_states** dataset)

```{r include=TRUE}
#creating the base outline of zion in it's appropriate location
map_zion <- tm_shape(zion) +  
  tm_borders(lwd = 2, col = "black")

#specifying a vector of colors for the palette 
zion_palette <- c('Water' = "#081baa",
                        'Developed' = "#656872",
                        'Barren' = "#cac091",
                        'Forest' = "#216f0c",
                        'Shrubland' = "#83a550",
                        'Herbaceous' = "#f5c8f9",
                        'Cultivated' = "#aa4908",
                        'Wetlands' = "#16ced4")

map_zion +
  tm_shape(nlcd) +
  tm_raster(alpha = 0.7,
            palette = zion_palette) +
  tm_layout(legend.outside = TRUE) +
  tm_compass(type = "4star", 
             position = c("right", "top")) +
  tm_scale_bar(breaks = c(0, 5, 10),
               position = c("left", "bottom"))
  

zion_palette_map <- map_zion +
  tm_shape(nlcd) +
  tm_raster(alpha = 0.7,
            palette = zion_palette) +
  tm_layout(legend.outside = TRUE) +
  tm_compass(type = "4star",
             position = c("right", "top")) +
  tm_scale_bar(breaks = c(0, 5, 10),
               position = c("left", "bottom"))

```


```{r}
## Filter to Utah
utah <- us_states %>%
  filter(NAME == "Utah")

utah_region <- st_bbox(zion, #creates a box around zion
                       crs = st_crs(nlcd)) %>%  #specify coordinate reference system
  st_as_sfc() 

zion_inset_map <- tm_shape(nlcd) + 
  tm_raster(style = "cont", palette = zion_palette, legend.show = TRUE) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE) +
  tm_shape(zion) +
  tm_borders(lwd = 2,
             col = "red")

utah_map <- tm_shape(utah) + tm_polygons() +
  tm_shape(utah_region) + tm_borders(lwd = 2, col = "red")

zion_inset_map
print(utah_map, vp = viewport(0.8, 0.2, width = 0.3, height = 0.3))
```


## question 5

##### 20 points

Create an interactive map of the world (10)\
- include a legend (5)\
- change the color palette (5)\
- bonus: use **leaflet** instead of **tmap** (2)

```{r include=TRUE}
tmap_mode("view") #changes the mode to be interactive!

tm_shape(world) +
  tm_fill(col = 'lifeExp',
          palette = "viridis") +
  tm_borders() +
  tm_layout(legend.outside =  TRUE)

```

```{r}
#LEAFLET

library(maptools)
library(leaflet)
library(rgeos)

pal <- colorNumeric("viridis",
                    domain = world$lifeExp)

labels <- sprintf(    #get the labels to work interactively, although I'm not sure why it's saying "people per square mile" rather than "life expectancy." 
  "<strong>%s</strong><br/>%g people / mi<sup>2</sup>",
  world$name_long, world$lifeExp
) %>% lapply(htmltools::HTML)   


leaflet(world) %>%
  addPolygons(
    fillColor = ~pal(world$lifeExp), 
    stroke = TRUE, 
    color = 'Steelblue', 
    weight = 1.5,
    highlightOptions = highlightOptions(
    weight = 5),
    label = labels) %>%
  addLegend(pal = pal, values = ~lifeExp, opacity = 0.7, title = NULL,
  position = "bottomright")
```


## bonus question

##### 5 points

Create THE WORST map! AKA a map that breaks all of the rules of legibility/clarity, yet somehow still passes for a map. We'll vote on the best worst map (worst worst map?) in class.

```{r include=TRUE}
##BAD MAP
tmap_mode("plot")

tm_shape(world) + 
  tm_fill(col = "name_long") +
  tm_borders(lwd = 10) +
  tm_layout(title = "cOoL GrAPh Yo",
            title.size = 2,
            title.bg.color = "green")

```